Compositional Quantitative Resource Analysis of C Programs

### 2014-06-17 paper/main.tex
start working on the logic
generalize assert rule
more rules
more work on section 2

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful practice.  That is why we developed a
new automatic amortized analysis for deriving quantitative Hoare
triples for common C programs.

@@

We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}p{\Phi'}$ in the quantitative
logic but enables inference using and LP solver.  This idea is best
explained by example.  Assume for the remainder of this section that
we are interested in bounding the number of \emph{ticks} in a program.
In other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The following example can then be bounded by $|[x,y]|$.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(1);}
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we exploit the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the update version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement $tick(1)$.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
assertions in $\Gamma$ are of the form XXXX.  They are only used to
determine at the assignment if we can extract constant potential to
pay for future cost.  It is important to note that the reasoning the
assertions in $\Gamma$ are very basic.  We do not perform any fixpoint
computations or invariants.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add \code{tick(1)} to
every back edge (loops and function calls) of the program.  However, a
potential function is more flexible.  For example, we can use a
potential function to prove that the following program does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);}
\end{lstlisting}
Similarly we can prove that the following program can be bounded by
$10|[x,y]|$.  In both cases, we reason exactly like in the first
version of the while loop to prove the bound.  Of course, such loops
with different notations can be seamlessly combined in a larger
program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);}
\end{lstlisting}

To understand how the analysis works, it is instructive to look at some
more involved examples.

@@

\and \Rule{S:FunCall}
{ \Delta~f = (x_1, \dots, x_n, S) \\
  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
}

@@

@@

\cleardoublepage
\appendix

>>>>

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful in practice.  That is why we developed a
new automatic amortized analysis for deriving quantitative Hoare
triples for common C programs.

@@

We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.  This idea is best
explained by example.  Assume for the remainder of this section that
we are interested in bounding the number of \emph{ticks} in a program.
In other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The cost of the following example can then be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we exploit the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x<y) {
  ${\color{blue}%
  x{<}y; \; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+1; 
  ${\color{blue}%
  \dashv x{\leq}y; \, 1 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv x{\leq}y; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x{\geq}y; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Derivation of a bound on the number of ticks for Example~1.}
  \label{fig:ex1}
\end{figure}
More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between.  It
is used to determine at the assignment if we can extract constant
potential to pay for future cost.  It is important to note that the
reasoning the assertions in $\Gamma$ are very basic.  We do not
perform any fixpoint computations or invariants.  \pref{fig:ex1} shows
the derivation for Example 1.  Note that the logical assertions in
$\Gamma$ are only used in the assignment \code{x=x+1}.

It would of course be possible to keep track more sophisticated
assertions in the logical context $\Gamma$.  However, this simple form
is sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$, which we are
interested in.  Nevertheless, this logical part of our analysis system
is orthogonal to the quantitative part and can be easily extended if
necessary.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at some
more involved examples.

XXX add up and down XXX

\begin{figure*}[t!]
 \setlength{\progwidth}{.22\linewidth}
  \centering
  \begin{minipage}[b]{.20\linewidth}
    \begin{center}
   \begin{lstlisting}
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y) 
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y = y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x = x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}

$|[x, n]| + |[y, m]|$
\\[.7\baselineskip]
      {\bf speed\_1}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}

while (x<n) {
  ${\color{blue}%
  x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
  if (z>x)
    ${\color{blue}%
    x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
    x=x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
  else
    ${\color{blue}%
    z{\leq}x, x{<}n; \, |[x, n]| {+} |[z, n]| \vdash }$
    z=z+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[z, n]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[z, n]| }$
  }
   \end{lstlisting}

$|[x, n]| + |[z, n]|$
\\[.7\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (z-y>0) {
    y = y+1;
    tick(3);
  }
  while (y>9) {
    y=y-10;
    tick(1);
  }
   \end{lstlisting}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100 && *){
    y=y-100;
    tick(5);
  }
  tick(9);
}
   \end{lstlisting}

$59|[n,0]| + 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning.}
  \label{fig:ex_list}
\end{figure*}

@@

\and \Rule{S:Call}
{ \delta~f = (x_1, \dots, x_n, S) \\
  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
}

@@

\begin{figure*}[t]
\begin{mathpar}

\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}

\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}

\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}

\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,\sem{e}_\state }{ \code{return}~e }{ Q }
}

\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}

\and
\Rule{L:Call}
{
  \Delta f = \forall z.(P_f\,z, Q_f\,z) \\
  P \models P_f \, y \, (\sem{e_1}_\state,\dots,\sem{e_n}_\state) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[x \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { x \gets f(e_1, \dots, e_n) }
  { Q }
}

\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}

\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}

\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}

\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}

\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z.(P_f\,z, Q_f\,z)
  \rightarrow
  \forall y.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

\end{mathpar}

\caption{Rules of the Quantitative Hoare Logic}
\end{figure*}

@@

\clearpage
\appendix
